---
title: "biochar"
output: html_notebook
---

This notebook contains the code to reproduce Supplementary Figure 9

Load libraries
```{r}
source("../../General/general_functions.R")
library(ggdendro)
library(vegan)
library(tidyverse)
```

Load and format data
```{r}
map <- readRDS("../Data/full_map.RDS")
v.map <- filter(map, Extraction == "VFD" & Keep)
b.map <- filter(map, Extraction == "BULK" & Keep)

# vOTU profiles from viromes
vfd.otu <- readRDS("../Data/virome_tpm_75_mtx.RDS")
v.vfd.otu <- vfd.otu[,colnames(vfd.otu) %in% v.map$SampleID]
v.vfd.otu <- v.vfd.otu[,match(colnames(v.vfd.otu), v.map$SampleID)] 
v.vfd.otu.ra <- rel_ab(v.vfd.otu)

# 16S profiles from Total MG
rrna.otu <- readRDS("../Data/rrna_otu_filt.RDS")
b.rrna.otu <- rrna.otu[,colnames(rrna.otu) %in% b.map$SampleID]
b.rrna.otu <- b.rrna.otu[,match(colnames(b.rrna.otu), b.map$SampleID)]
b.rrna.otu.ra <- rel_ab(b.rrna.otu)
```


Calculate the Bray-Curtis dissimilarities on Hellinger-transformed relative abundances
```{r}
v.vfd.dist <- vegan::vegdist(vegan::decostand(t(v.vfd.otu.ra), method = "hellinger"), method = "bray") %>% as.matrix()

b.rrna.dist <- vegan::vegdist(vegan::decostand(t(b.rrna.otu.ra), method = "hellinger"), method = "bray") %>% as.matrix()
```

Since the effect of biochar on viral communities is only detected after accounting for collection time point ("Month" variable) and W-E gradient ("Column" variable), we need to run a CAP that tests for the effect of biochar while controlling for these other two variables.
```{r}
# Run the CAP and extract the values needed  for plotting
v.cap <- vegan::capscale(as.dist(v.vfd.dist) ~ Biochar + Condition(Month + Column), data = v.map, add = T, na.action = na.omit)
v.cap.axes <- cap_axes(v.cap, v.map)
v.cap.eig <- cap_eigval(v.cap, v.map)

#Plot
v.bch.cap.p <- v.cap.axes %>% 
  mutate(BiocharFull = fct_relevel(BiocharFull, "No Biochar", after = Inf)) %>% 
  ggplot(aes(CAP1, CAP2, color = BiocharFull, shape = Month)) +
  geom_point(size = 4, stroke = 1) +
  xlab(paste("CAP1 (", v.cap.eig$Eigval[1], "%)", sep = "")) +
  ylab(paste("CAP2 (", v.cap.eig$Eigval[2], "%)", sep = "")) +
  scale_color_brewer(name = "Biochar", palette = "Dark2") +
  theme_light() +
  theme(text = element_text(size = 15))

v.bch.cap.p
```

While the effect of biochar on bacterial communities is not significant, we still ran a CAP controlling for time point (the only variable significantly structuring the 16S data) to see if there was any discernable patterns
```{r}
# Run the CAP and extract the values needed  for plotting
b.cap <- vegan::capscale(as.dist(b.rrna.dist) ~ Biochar + Condition(Month), data = b.map, add = T, na.action = na.omit)
b.cap.axes <- cap_axes(b.cap, b.map)
b.cap.eig <- cap_eigval(b.cap, b.map)

#Plot
b.bch.cap.p <- b.cap.axes %>% 
  mutate(BiocharFull = fct_relevel(BiocharFull, "No Biochar", after = Inf)) %>% 
  ggplot(aes(CAP1, CAP2, color = BiocharFull, shape = Month)) +
  geom_point(size = 4, stroke = 1) +
  xlab(paste("CAP1 (", b.cap.eig$Eigval[1], "%)", sep = "")) +
  ylab(paste("CAP2 (", b.cap.eig$Eigval[2], "%)", sep = "")) +
  scale_color_brewer(name = "Biochar", palette = "Dark2") +
  theme_light() +
  theme(text = element_text(size = 15))

b.bch.cap.p
```

Identify the vOTUs differentially abundant across biochar treatments and inspect their abundance patterns
```{r}
# Load DESeq results and format
sig.bch <- readRDS("../Data/v_deseq_bch.RDS") %>% filter(p.adjusted < 0.05)
tidy.otu <- v.vfd.otu.ra %>% 
  tidy_otu()

# Generate a matrix displaying the mean z-transformed relative abundances of differentially abundant vOTUs across treatments
bch.mtx <- tidy.otu %>% 
  filter(OTU_ID %in% sig.bch$OTU_ID) %>% 
  group_by(OTU_ID) %>% 
  mutate(zValue = (Count - mean(Count))/sd(Count)) %>% 
  inner_join(map, by = "SampleID") %>% 
  group_by(Biochar2, OTU_ID) %>% 
  summarise(MeanZS = mean(zValue)) %>% 
  spread(key = OTU_ID, value = MeanZS) %>% 
  as.data.frame()
  
row.names(bch.mtx) <- bch.mtx$Biochar2
bch.mtx <- bch.mtx[,-1]
bch.mtx <- as.matrix(bch.mtx)

# Calculate distances 
bch.dist <- dist(bch.mtx)
otu.dist <- dist(t(bch.mtx))

# Perform hierarchical clustering
bch.dd <- as.dendrogram(hclust(as.dist(bch.dist), method = "complete"))
bch.ddata_x <- dendro_data(bch.dd)
bch.labs <- label(bch.ddata_x) %>% 
  rename("Biochar" = "label") %>% 
  rename("BiocharOrder" = "x")

otu.dd <- as.dendrogram(hclust(as.dist(otu.dist), method = "complete"))
otu.ddata_x <- dendro_data(otu.dd)
otu.labs <- label(otu.ddata_x) %>% 
  rename("OTU_ID" = "label") %>% 
  rename("OTUOrder" = "x")

# Plot a heatmap displaying the relative abundances patterns across biochar treatments
bch.hm.p <- tidy.otu %>% 
  filter(OTU_ID %in% sig.bch$OTU_ID) %>% 
  group_by(OTU_ID) %>% 
  mutate(zValue = (Count - mean(Count))/sd(Count)) %>% 
  inner_join(map, by = "SampleID") %>% 
  group_by(Biochar2, OTU_ID) %>% 
  summarise(MeanZS = mean(zValue)) %>% 
  rename("Biochar" = "Biochar2") %>% 
  inner_join(select(bch.labs, Biochar, BiocharOrder), by = "Biochar") %>% 
  inner_join(select(otu.labs, OTU_ID, OTUOrder), by = "OTU_ID") %>% 
  ggplot(aes(reorder(Biochar, BiocharOrder), reorder(OTU_ID, OTUOrder), fill = MeanZS)) +
  geom_tile() +
  ylab("vOTU") +
  scale_fill_distiller(name = "Mean Rel. Abund.\n(z-score)",palette = "Greys",
                       guide = guide_colorbar(title.hjust = 0,
                                              title.position = "top",
                                              label.theme = element_text(angle = 90, hjust = 1))) +
  theme_minimal() +
  theme(text = element_text(size = 15),
        axis.text = element_blank(),
        axis.title.x = element_blank(),
        legend.position = "bottom")


# Plot the hierarchical clustering dendogram
bch.dd.p <- ggplot(segment(bch.ddata_x)) +
  geom_segment(aes(x=x, y=y, xend=xend, yend=yend))

bch.dd.labs <- label(bch.ddata_x) %>% 
  mutate(label = fct_relevel(label, "AS", "CS", "PN", "NB"))

bch.dendo.p <- bch.dd.p + 
  geom_point(data=bch.dd.labs,
             aes(x=x, y=1, color = label), size = 4, shape = 15) +
  scale_color_brewer(name = "Biochar", palette = "Dark2") +
  xlim(0.5, 4.5) +
  theme_classic() +
  theme(text = element_text(size = 15),
        #legend.text = element_text(size = 15),
        #legend.title = element_blank(),
        axis.text = element_blank(),
        axis.title = element_blank(),
        axis.line = element_blank(),
        axis.ticks = element_blank(),
        legend.position = "none")
```

Put all plots together into one figure
```{r}
left <- cowplot::plot_grid(v.bch.cap.p + theme(legend.position = "none"), 
                           b.bch.cap.p + theme(legend.position = "none"), 
                           nrow = 2,
                           labels = c("A","B"), 
                           label_size = 20)

right <- cowplot::plot_grid(bch.dendo.p, 
                            bch.hm.p + theme(legend.position = "none"), 
                            ncol = 1, 
                            align = "v", axis = "r", 
                            rel_heights = c(1,4),
                            labels = c("C"), 
                            label_size = 20)

bch.legend <- cowplot::plot_grid(cowplot::get_legend(v.bch.cap.p), cowplot::get_legend(bch.hm.p), ncol = 1, rel_heights = c(5,2))

cowplot::plot_grid(left, right, bch.legend, nrow = 1, rel_widths = c(2,1,1))
```

