---
title: "permanova"
output: html_notebook
---

This notebook contains the code to reproduce Supplementary Tables 4 & 5

Load libraries
```{r}
source("../../General/general_functions.R")
library(vegan)
library(tidyverse)
```

Load data and filter appropriate sets
```{r}
map <- readRDS("../Data/full_map.RDS")
v.map <- filter(map, Extraction == "VFD" & Keep)
b.map <- filter(map, Extraction == "BULK" & Keep)

vfd.otu <- readRDS("../Data/virome_tpm_75_mtx.RDS")
v.vfd.otu <- vfd.otu[,colnames(vfd.otu) %in% v.map$SampleID]
v.vfd.otu <- v.vfd.otu[,match(colnames(v.vfd.otu), v.map$SampleID)] 
v.vfd.otu.ra <- rel_ab(v.vfd.otu)

rrna.otu <- readRDS("../Data/rrna_otu_filt.RDS")
b.rrna.otu <- rrna.otu[,colnames(rrna.otu) %in% b.map$SampleID]
b.rrna.otu <- b.rrna.otu[,match(colnames(b.rrna.otu), b.map$SampleID)]
b.rrna.otu.ra <- rel_ab(b.rrna.otu)
```

Calculate the Bray-Curtis dissimilarites on Hellinger-transformed relative abundances
```{r}
v.vfd.dist <- vegan::vegdist(vegan::decostand(t(v.vfd.otu.ra), method = "hellinger"), method = "bray") %>% 
  as.matrix()
b.rrna.dist <- vegan::vegdist(vegan::decostand(t(b.rrna.otu.ra), method = "hellinger"), method = "bray") %>% 
  as.matrix()
```

Since nitrogen applications were started after the first time point, we can only test the effect of nitrogen fertilizer concentration on the August subset.
```{r}
aug.v.map <- filter(v.map, Month == "Aug")
aug.b.map <- filter(b.map, Month == "Aug")

aug.v.vfd.dist <- v.vfd.dist[row.names(v.vfd.dist) %in% aug.v.map$SampleID, colnames(v.vfd.dist) %in% aug.v.map$SampleID]
aug.b.rrna.dist <- b.rrna.dist[row.names(b.rrna.dist) %in% aug.b.map$SampleID, colnames(b.rrna.dist) %in% aug.b.map$SampleID]
```

Perform individual PERMANOVAs testing the effect of each experimental factor (collection time point, biochar treatment, nitrogen concentration) and spatial variable (W-E position along block columns and N-S position along block rows) on community composition. Note that the spatial variables are coded as continuous variables.  
```{r}
# PERMANOVA for vOTU profiles in viromes
v.pm.month <- vegan::adonis(as.dist(v.vfd.dist) ~ Month,  data = v.map)$aov.tab %>% 
  as.data.frame() %>% 
  mutate(Term = row.names(.), Extraction = "VIR", Set = "vOTU", Formula = "dist ~ Month")

v.pm.bch <- vegan::adonis(as.dist(v.vfd.dist) ~ Biochar,  data = v.map)$aov.tab %>% 
  as.data.frame() %>% 
  mutate(Term = row.names(.), Extraction = "VIR", Set = "vOTU", Formula = "dist ~ Biochar")

v.pm.col <- vegan::adonis(as.dist(v.vfd.dist) ~ Column,  data = v.map)$aov.tab %>% 
  as.data.frame() %>% 
  mutate(Term = row.names(.), Extraction = "VIR", Set = "vOTU", Formula = "dist ~ Column")

v.pm.row <- vegan::adonis(as.dist(v.vfd.dist) ~ Row,  data = v.map)$aov.tab %>% 
  as.data.frame() %>% 
  mutate(Term = row.names(.), Extraction = "VIR", Set = "vOTU", Formula = "dist ~ Row")

v.pm.nit <- vegan::adonis(as.dist(aug.v.vfd.dist) ~ Nitrogen,  data = aug.v.map)$aov.tab %>% 
  as.data.frame() %>% 
  mutate(Term = row.names(.), Extraction = "VIR", Set = "vOTU", Formula = "dist ~ Nitrogen")


# PERMANOVA for 16SOTU profiles in total MGs
b.pm.month <- vegan::adonis(as.dist(b.rrna.dist) ~ Month,  data = b.map)$aov.tab %>% 
  as.data.frame() %>% 
  mutate(Term = row.names(.), Extraction = "TMG", Set = "16SOTU", Formula = "dist ~ Month")

b.pm.bch <- vegan::adonis(as.dist(b.rrna.dist) ~ Biochar,  data = b.map)$aov.tab %>% 
  as.data.frame() %>% 
  mutate(Term = row.names(.), Extraction = "TMG", Set = "16SOTU", Formula = "dist ~ Biochar")

b.pm.col <- vegan::adonis(as.dist(b.rrna.dist) ~ Column,  data = b.map)$aov.tab %>% 
  as.data.frame() %>% 
  mutate(Term = row.names(.), Extraction = "TMG", Set = "16SOTU", Formula = "dist ~ Column")

b.pm.row <- vegan::adonis(as.dist(b.rrna.dist) ~ Row,  data = b.map)$aov.tab %>% 
  as.data.frame() %>% 
  mutate(Term = row.names(.), Extraction = "TMG", Set = "16SOTU", Formula = "dist ~ Row")

b.pm.nit <- vegan::adonis(as.dist(aug.b.rrna.dist) ~ Nitrogen,  data = aug.b.map)$aov.tab %>% 
  as.data.frame() %>% 
  mutate(Term = row.names(.), Extraction = "TMG", Set = "16SOTU", Formula = "dist ~ Nitrogen")

pmanova.single <- rbind(v.pm.month, v.pm.bch, v.pm.col, v.pm.row, v.pm.nit,
                        b.pm.month, b.pm.bch, b.pm.col, b.pm.row, b.pm.nit)

# Save supplementary table
write.table(pmanova.single, "../Tables/pmanova_single.tsv", sep = "\t", quote = F, row.names = F)
```

Given that collection time point and W-E position had a significant effect on viral community composition, we performed new sets of PERMANOVAs testing biochar treatment or nitrogen fertilization regime while accounting for collection time point and/or W-E position
```{r}
# PERMANOVA for vOTU profiles in viromes
v.pm.bch.month <- vegan::adonis(as.dist(v.vfd.dist) ~ Biochar + Month,  data = v.map)$aov.tab %>% 
  as.data.frame() %>% 
  mutate(Term = row.names(.), Extraction = "VIR", Set = "vOTU", Formula = "dist ~ Biochar + Month")

v.pm.bch.col <- vegan::adonis(as.dist(v.vfd.dist) ~ Biochar + Column,  data = v.map)$aov.tab %>% 
  as.data.frame() %>% 
  mutate(Term = row.names(.), Extraction = "VIR", Set = "vOTU", Formula = "dist ~ Biochar + Column")

v.pm.bch.month.col <- vegan::adonis(as.dist(v.vfd.dist) ~ Biochar + Month + Column,  data = v.map)$aov.tab %>% 
  as.data.frame() %>% 
  mutate(Term = row.names(.), Extraction = "VIR", Set = "vOTU", Formula = "dist ~ Biochar + Month + Column")

v.pm.nit.col <- vegan::adonis(as.dist(aug.v.vfd.dist) ~ Nitrogen + Column,  data = aug.v.map)$aov.tab %>% 
  as.data.frame() %>% 
  mutate(Term = row.names(.), Extraction = "VIR", Set = "vOTU", Formula = "dist ~ Nitrogen + Column")

# PERMANOVA for 16S OTU profiles in total MGs
b.pm.bch.month <- vegan::adonis(as.dist(b.rrna.dist) ~ Biochar + Month,  data = b.map)$aov.tab %>% 
  as.data.frame() %>% 
  mutate(Term = row.names(.), Extraction = "TMG", Set = "16SOTU", Formula = "dist ~ Biochar + Month")

b.pm.bch.col <- vegan::adonis(as.dist(b.rrna.dist) ~ Biochar + Column,  data = b.map)$aov.tab %>% 
  as.data.frame() %>% 
  mutate(Term = row.names(.), Extraction = "TMG", Set = "16SOTU", Formula = "dist ~ Biochar + Column")

b.pm.bch.month.col <- vegan::adonis(as.dist(b.rrna.dist) ~ Biochar + Month + Column,  data = b.map)$aov.tab %>% 
  as.data.frame() %>% 
  mutate(Term = row.names(.), Extraction = "TMG", Set = "16SOTU", Formula = "dist ~ Biochar + Month + Column")

b.pm.nit.col <- vegan::adonis(as.dist(aug.b.rrna.dist) ~ Nitrogen + Column,  data = aug.b.map)$aov.tab %>% 
  as.data.frame() %>% 
  mutate(Term = row.names(.), Extraction = "TMG", Set = "16SOTU", Formula = "dist ~ Nitrogen + Column")

pmanova.multiple <- rbind(v.pm.bch.month, v.pm.bch.col, v.pm.bch.month.col, v.pm.nit.col,
                          b.pm.bch.month, b.pm.bch.col, b.pm.bch.month.col, b.pm.nit.col)

# Save supplementary table
write.table(pmanova.multiple, "../Tables/pmanova_multiple.tsv", sep = "\t", quote = F, row.names = F)
```
