---
title: "spatial"
output: html_notebook
---

This notebook contains the code to reproduce Figure 5 and Supplementary Figures 7 & 8

Load libraries
```{r}
source("../../General/general_functions.R")
library(vegan)
library(tidyverse)
```

Load data and reformat
```{r}
map <- readRDS("../Data/full_map.RDS")
v.map <- filter(map, Extraction == "VFD" & Keep)
b.map <- filter(map, Extraction == "BULK" & Keep)

plot.des <- readRDS("../Data/plot_design.RDS")
otu <- readRDS("../Data/virome_tpm_75_mtx.RDS")

v.otu <- otu[,colnames(otu) %in% v.map$SampleID]
v.otu.ra <- rel_ab(v.otu)

b.otu <- readRDS("../Data/rrna_otu_filt.RDS")
b.otu <- b.otu[,colnames(b.otu) %in% b.map$SampleID]
b.otu.ra <- rel_ab(b.otu)

spatial.pal <- RColorBrewer::brewer.pal(4, "Purples") 
```

Perform unconstrained PCoA
```{r}
# Calculate Bray-Curtis dissimilarities on Hellinger-transformed relative abundances
v.dist <- vegan::vegdist(vegan::decostand(t(v.otu.ra), method = "hellinger"), method = "bray")

# Get PCoA values
v.pcoa.axes <- pcoa_axes(as.matrix(v.dist), v.map)
v.pcoa.eig <- pcoa_eigval(as.matrix(v.dist), v.map)

# Plot
v.pcoa.p <- v.pcoa.axes %>% 
  mutate(Biochar = strtrim(Sample, 2),
         Nitrogen = str_extract(Sample, "L|H")) %>% 
  ggplot(aes(Axis.1, Axis.2)) +
  geom_point(aes(fill = as.factor(Position_x), shape = Month), 
             size = 4, 
             color = "gray25", 
             stroke = 1) +
  geom_line(aes(group = Sample), color = "gray25", linetype = 2) +
  scale_color_brewer(name = "", palette = "Paired") +
  scale_shape_manual(name = "", values = c(21,24)) +
  scale_fill_manual(values = spatial.pal, guide = F) +
  scale_color_manual(values = spatial.pal, guide = F) +
  xlab(paste("PCo1 (", v.pcoa.eig$Eigval[1], "%)", sep = "")) +
  ylab(paste("PCo2 (", v.pcoa.eig$Eigval[2], "%)", sep = "")) +
  theme_light() +
  theme(text = element_text(size = 15),
        legend.position = "top")

v.pcoa.p
```

Plot the same ordination colored by biochar and nitrogen ammendments to show the lack of patterns across the first and second axes
```{r}
# Color-coded by biochar treatment
bch.p <- v.pcoa.axes %>% 
  mutate(BiocharFull = fct_relevel(BiocharFull, "No Biochar", after = Inf)) %>% 
  ggplot(aes(Axis.1, Axis.2)) +
  geom_point(aes(fill = BiocharFull, shape = Month), 
             size = 4, 
             stroke = 1) +
  scale_shape_manual(name = "Month", values = c(21,24)) +
  scale_fill_brewer(name = "Biochar", palette = "Dark2") +
  xlab(paste("PCo1 (", v.pcoa.eig$Eigval[1], "%)", sep = "")) +
  ylab(paste("PCo2 (", v.pcoa.eig$Eigval[2], "%)", sep = "")) +
  theme_light() +
  theme(text = element_text(size = 15),
        legend.position = "right")

# Color-coded by nitrogen fertilizer concentration
nit.p <- v.pcoa.axes %>% 
  mutate(Nitrogen = case_when(Month == "Apr" ~ "Pre-Fertigation",
                              Nitrogen == "H" ~ "225 lbs N/acre",
                              Nitrogen == "L" ~ "150 lbs N/acre")) %>% 
  mutate(Nitrogen = fct_relevel(Nitrogen, "Pre-Fertigation")) %>% 
  ggplot(aes(Axis.1, Axis.2)) +
  geom_point(aes(fill = Nitrogen, shape = Month), 
             size = 4, 
             stroke = 1) +
  scale_shape_manual(name = "Month", values = c(21,24)) +
  scale_fill_manual(values = c("white", "gray75", "gray25")) +
  xlab(paste("PCo1 (", v.pcoa.eig$Eigval[1], "%)", sep = "")) +
  ylab(paste("PCo2 (", v.pcoa.eig$Eigval[2], "%)", sep = "")) +
  theme_light() +
  theme(text = element_text(size = 15),
        legend.position = "right")

supp.votu <- cowplot::plot_grid(bch.p, nit.p, nrow = 2, align = "v", labels = c("A", "B"), label_size = 20)
supp.votu
```
Plot the distribution of sampled plots in the field
```{r}
field.p <- plot.des %>% 
  ggplot(aes(Position_x_mts, Position_y_mts)) +
  geom_tile(aes(fill = as.factor(Column)), color = "black", size = 1) +
  geom_point(data = filter(plot.des, !is.na(Sample)), aes(color = Column), size = 3, shape = 4, stroke = 1) +
  scale_fill_manual(values = spatial.pal) +
  scale_color_distiller(palette = "Greys", direction = -1) +
  xlab("Distance (m)\nWest-East") +
  ylab("Distance (m)\nSouth-North") +
  coord_fixed() +
  theme_bw() +
  theme(text = element_text(size = 15),
        panel.border = element_blank(),
        legend.position = "none")

field.p
```

Calculate the pairwise W-E distances between plots
```{r}
tmp <- map %>% filter(!is.na(Sample)) %>% 
  group_by(Sample, Position_x_mts) %>% 
  count() %>% 
  select(-n) %>% 
  as.data.frame()

field.dist <- data.frame(FieldDist = tmp$Position_x_mts,
                         row.names = tmp$Sample) %>% 
  dist()

field.dist <- as.matrix(field.dist)

# Generate a data frame with W-E distances for plotting
field.tidy <- field.dist  %>%
  as.matrix() %>%
  as.data.frame() %>%
  mutate(Sample.x = row.names(.)) %>%
  gather(key = "Sample.y", value = "FieldDistance", -Sample.x) %>%
  filter(Sample.x != Sample.y)

```

Perform Mantel tests comparing viral community composition dissimilarities vs spatial distance (calculated within each month)
```{r}
v.map.apr <- filter(v.map, Month == "Apr")
v.map.aug <- filter(v.map, Month == "Aug")

v.dist <- as.matrix(v.dist)

# April samples
v.apr.dist <- v.dist[match(v.map.apr$SampleID,row.names(v.dist)),match(v.map.apr$SampleID,colnames(v.dist))]
row.names(v.apr.dist) <- str_split(row.names(v.apr.dist), "_", n = 2, simplify = T)[,1]
colnames(v.apr.dist) <- str_split(colnames(v.apr.dist), "_", n = 2, simplify = T)[,1]
v.apr.field.dist <- field.dist[match(row.names(v.apr.dist), row.names(field.dist)), match(colnames(v.apr.dist), colnames(field.dist))]
v.apr.mant <- vegan::mantel(as.dist(v.apr.dist), as.dist(v.apr.field.dist), permutations = 9999, method = "pearson") 
v.apr.mant

# August samples
v.aug.dist <- v.dist[match(v.map.aug$SampleID,row.names(v.dist)),match(v.map.aug$SampleID,colnames(v.dist))]
row.names(v.aug.dist) <- str_split(row.names(v.aug.dist), "_", n = 2, simplify = T)[,1]
colnames(v.aug.dist) <- str_split(colnames(v.aug.dist), "_", n = 2, simplify = T)[,1]
v.aug.field.dist <- field.dist[match(row.names(v.aug.dist), row.names(field.dist)), match(colnames(v.aug.dist), colnames(field.dist))]
v.aug.mant <- vegan::mantel(as.dist(v.aug.dist), as.dist(v.aug.field.dist), permutations = 9999, method = "pearson") 
v.aug.mant

# Generate a data frame with correlation values for plotting
v.mant.df <- data.frame(Month.x = c("Apr", "Aug"),
                      R = c(v.apr.mant$statistic,v.aug.mant$statistic),
                      P = c(v.apr.mant$signif, v.aug.mant$signif)) %>% 
  mutate(r = paste("r", signif(R, 3), sep = " == "), 
         Pval = paste("P", signif(P,2), sep = " == ")) %>% 
  mutate(Label = paste(Month.x, r, Pval, sep = "\n")) %>% 
  mutate(Label = str_replace_all(Label, "==", "="))
```

Plot the correlation between W-E distances and vOTU Bray-Curtis dissimilarities
```{r}
# Generate a data frame with Bray-Curtis dissimilarities for plotting
v.beta.tidy <- as.matrix(v.dist) %>%
  as.data.frame() %>%
  mutate(SampleID.x = row.names(.)) %>%
  gather(key = "SampleID.y", value = "BetaDistance", -SampleID.x) %>%
  inner_join(select(v.map, SampleID, Sample, Month), by = c("SampleID.x" = "SampleID")) %>%
  inner_join(select(v.map, SampleID, Sample, Month), by = c("SampleID.y" = "SampleID")) %>%
  filter(SampleID.x != SampleID.y) %>%
  filter(Month.x == Month.y)

v.pair.tidy <- inner_join(v.beta.tidy, field.tidy, by = c("Sample.x", "Sample.y"))

# Plot
v.cor.p <- v.pair.tidy %>% 
  ggplot(aes(FieldDistance, BetaDistance)) +
  geom_point(aes(color = Month.x)) +
  geom_smooth(aes(color = Month.x), method = "lm", se = F) +
  geom_label(data = filter(v.mant.df, Month.x == "Apr"), aes(0, 
                                 max(v.pair.tidy$BetaDistance), 
                                 label = Label, 
                                 fill = Month.x), 
            hjust = 0, 
            vjust = 1, 
            parse = F) +
  geom_label(data = filter(v.mant.df, Month.x == "Aug"), aes(5, 
                                 min(v.pair.tidy$BetaDistance), 
                                 label = Label, 
                                 fill = Month.x), 
            hjust = 0, 
            vjust = 0, 
            parse = F) +
  scale_color_brewer(name = "", palette = "Paired") +
  scale_fill_brewer(name = "", palette = "Paired") +
  ylab("vOTUs (Virome)\nPairwise B-C Dissimilarity") +
  xlab("Pairwise W-E Distance (m)") +
  theme_light() +
  theme(text = element_text(size = 15),
        legend.position = "none")

v.cor.p
```

Plot the abundance patterns of spatially structured vOTUs. 
```{r}
# Load DESeq2 results
sig.spatial <- read.table("../Tables/sig_vOTUs_spatial.tsv", sep = "\t", header = T)

# Get the relative abundances
v.tidy.otu <- v.otu.ra %>% 
  tidy_otu()

# Data frame to plot the W-E position for each sampled column
col.tmp <- v.tidy.otu %>% 
  inner_join(sig.spatial, by = "OTU_ID") %>% 
  inner_join(v.map, by = "SampleID") %>% 
  group_by(SampleID, SpatialEnrichment, TemporalEnrichment, Column, Month) %>% 
  summarise(AggRelAb = sum(Count)) %>% 
  group_by(SpatialEnrichment, Column, Month) %>% 
  count() %>% 
  ungroup() %>% 
  mutate(SpatialEnrichment = fct_relevel(SpatialEnrichment, "West-Enriched")) 

# Plot the aggregated relative abundances within a single column and identify which vOTUs were also affected by collection time point
bar.p <- v.tidy.otu %>% 
  inner_join(sig.spatial, by = "OTU_ID") %>% 
  inner_join(map, by = "SampleID") %>% 
  group_by(SampleID, SpatialEnrichment, TemporalEnrichment, Column, Month) %>% 
  summarise(AggRelAb = sum(Count)) %>% 
  group_by(SpatialEnrichment, TemporalEnrichment, Column, Month) %>% 
  summarise(MeanAggRelAb = mean(AggRelAb)/100) %>% 
  ungroup() %>% 
  mutate(SpatialEnrichment = fct_relevel(SpatialEnrichment, "West-Enriched")) %>% 
  ggplot(aes(Column, MeanAggRelAb)) +
  geom_bar(stat = "identity", position = "stack", aes(fill = TemporalEnrichment)) +
  geom_point(data = col.tmp, aes(y = 0, color = as.factor(Column), shape = Month), size = 4, stroke = 1) +
  scale_fill_manual(name = "Time Point\nEnrichment",values = c(RColorBrewer::brewer.pal(4, "Paired")[3:4], "gray")) +
  scale_color_manual(values = spatial.pal, guide = F) +
  scale_shape(guide = F) +
  ylab("Mean Rel. Ab.") +
  xlab("West-East Position") +
  facet_grid(. ~ SpatialEnrichment + Month) +
  theme_light() +
  theme(text=element_text(size = 15),
        axis.text.x = element_blank(),
        legend.position = "right")
  
bar.p
```

Plot it all together
```{r}
#841:611
a <- cowplot::plot_grid(v.pcoa.p, field.p, rel_widths = c(5,4),
                        labels = c("A", "B"), label_size = 20)

b <- cowplot::plot_grid(a, v.cor.p, nrow = 1, rel_widths = c(3,2), 
                        labels = c(NA, "C", "D"), label_size = 20)

cowplot::plot_grid(b, bar.p, nrow = 2, rel_heights = c(7,6),labels = c(NA, "D"), label_size = 20)
```

Do the same for the 16S rRNA gene communities extracted from Total MG
Plot PCoA
```{r}
b.dist <- vegan::vegdist(vegan::decostand(t(b.otu.ra), method = "hellinger"), method = "bray")

b.pcoa.axes <- pcoa_axes(as.matrix(b.dist), b.map)
b.pcoa.eig <- pcoa_eigval(as.matrix(b.dist), b.map)

b.pcoa.p <- b.pcoa.axes %>% 
  ggplot(aes(Axis.1, Axis.2)) +
  xlab(paste("PCo1 (", b.pcoa.eig$Eigval[1], "%)", sep = "")) +
  ylab(paste("PCo2 (", b.pcoa.eig$Eigval[2], "%)", sep = "")) +
  geom_point(aes(color = Month, shape = Month), size = 4, stroke = 1) +
  scale_shape_manual(name = "16S OTUs\n(Total MG)", values = c(16,17)) +
  scale_color_manual(name = "16S OTUs\n(Total MG)", values = RColorBrewer::brewer.pal(6, "Paired")[c(5,6)]) +
  theme_bw() +
  theme(text = element_text(size = 15),
        legend.position = "left")
b.pcoa.p
```

Perform Mantel tests comparing 16S community composition dissimilarities vs spatial distance (calculated within each month)
```{r}
b.map.apr <- filter(b.map, Month == "Apr")
b.map.aug <- filter(b.map, Month == "Aug")

b.dist <- as.matrix(b.dist)

b.apr.dist <- b.dist[match(b.map.apr$SampleID,row.names(b.dist)),match(b.map.apr$SampleID,colnames(b.dist))]
row.names(b.apr.dist) <- str_split(row.names(b.apr.dist), "_", n = 2, simplify = T)[,1]
colnames(b.apr.dist) <- str_split(colnames(b.apr.dist), "_", n = 2, simplify = T)[,1]
b.apr.field.dist <- field.dist[match(row.names(b.apr.dist), row.names(field.dist)), match(colnames(b.apr.dist), colnames(field.dist))]
b.apr.mant <- vegan::mantel(as.dist(b.apr.dist), as.dist(b.apr.field.dist), permutations = 9999, method = "pearson") 
b.apr.mant

b.aug.dist <- b.dist[match(b.map.aug$SampleID,row.names(b.dist)),match(b.map.aug$SampleID,colnames(b.dist))]
row.names(b.aug.dist) <- str_split(row.names(b.aug.dist), "_", n = 2, simplify = T)[,1]
colnames(b.aug.dist) <- str_split(colnames(b.aug.dist), "_", n = 2, simplify = T)[,1]
b.aug.field.dist <- field.dist[match(row.names(b.aug.dist), row.names(field.dist)), match(colnames(b.aug.dist), colnames(field.dist))]
b.aug.mant <- vegan::mantel(as.dist(b.aug.dist), as.dist(b.aug.field.dist), permutations = 9999, method = "pearson") 
b.aug.mant

b.mant.df <- data.frame(Month.x = c("Apr", "Aug"),
                      R = c(b.apr.mant$statistic,b.aug.mant$statistic),
                      P = c(b.apr.mant$signif, b.aug.mant$signif)) %>% 
  mutate(r = paste("r", signif(R, 3), sep = " == "), 
         Pval = paste("P", signif(P,2), sep = " == ")) %>% 
  mutate(Label = paste(Month.x, r, Pval, sep = "\n")) %>% 
  mutate(Label = str_replace_all(Label, "==", "="))
```
Plot the correlation between W-E distances and 16S OTU Bray-Curtis dissimilarities
```{r}
b.beta.tidy <- as.matrix(b.dist) %>%
  as.data.frame() %>%
  mutate(SampleID.x = row.names(.)) %>%
  gather(key = "SampleID.y", value = "BetaDistance", -SampleID.x) %>%
  inner_join(select(b.map, SampleID, Sample, Month), by = c("SampleID.x" = "SampleID")) %>%
  inner_join(select(b.map, SampleID, Sample, Month), by = c("SampleID.y" = "SampleID")) %>%
  filter(SampleID.x != SampleID.y) %>%
  filter(Month.x == Month.y)

b.pair.tidy <- inner_join(b.beta.tidy, field.tidy, by = c("Sample.x", "Sample.y"))

b.cor.p <- b.pair.tidy %>% 
  ggplot(aes(FieldDistance, BetaDistance)) +
  geom_point(aes(color = Month.x)) +
  geom_smooth(aes(color = Month.x), method = "lm", se = F) +
  geom_label(data = filter(b.mant.df, Month.x == "Apr"), aes(0, 
                                 max(b.pair.tidy$BetaDistance), 
                                 label = Label, 
                                 fill = Month.x), 
            hjust = 0, 
            vjust = 1, 
            parse = F) +
  geom_label(data = filter(b.mant.df, Month.x == "Aug"), aes(0, 
                                 min(b.pair.tidy$BetaDistance), 
                                 label = Label, 
                                 fill = Month.x), 
            hjust = 0, 
            vjust = 0, 
            parse = F) +
  scale_color_manual(values = RColorBrewer::brewer.pal(6, "Paired")[c(5,6)]) +
  scale_fill_manual(values = RColorBrewer::brewer.pal(6, "Paired")[c(5,6)]) +
  ylab("16S OTUs (Total MG)\nPairwise B-C Dissimilarity") +
  xlab("Pairwise W-E Distance (m)") +
  theme_bw() +
  theme(text = element_text(size = 15),
        legend.position = "none")

b.cor.p
```
Put it all together
```{r}
supp.16S <- cowplot::plot_grid(b.pcoa.p, b.cor.p,
                   nrow = 1,
                   align = "h",
                   axis = "l",
                   rel_widths = c(5,4),
                   labels = c("A", "B"), label_size = 20)

supp.16S
```


