---
title: "temporal"
output: html_notebook
---

This notebook contains the code to reproduce Figure 4A-B and Supplementary Figure 6

Load libraries
```{r}
source("../../General/general_functions.R")
library(dendextend)
library(ggdendro)
library(tidyverse)
```

Load and format data
```{r}
map <- readRDS("../Data/full_map.RDS")
v.map <- filter(map, Extraction == "VFD" & Keep) %>% 
  mutate(Group = paste(Sample, Month, sep = "."))

b.map <- filter(map, Extraction == "BULK" & Keep) %>% 
  mutate(Group = paste(Sample, Month, sep = "."))

v.otu <- readRDS("../Data/virome_tpm_75_mtx.RDS")
v.otu <- v.otu[,colnames(v.otu) %in% v.map$SampleID]
v.otu.ra <- rel_ab(v.otu)
v.otu.tidy <- tidy_otu(v.otu.ra)

b.otu <- readRDS("../Data/rrna_otu_filt.RDS")
b.otu <- b.otu[,colnames(b.otu) %in% b.map$SampleID]
b.otu.ra <- rel_ab(b.otu)
b.otu.tidy <- tidy_otu(b.otu.ra)
```

For each subset (vOTUs in viromes and 16S OTUs in Total MGs), calculate Bray-Curtis dissimilarities on Hellinger-transformed relative abundances.
```{r}
v.dist <- vegan::vegdist(vegan::decostand(t(v.otu.ra), method = "hellinger"), method = "bray") %>% as.matrix()
v.map <- v.map[match(row.names(v.dist), v.map$SampleID),]
row.names(v.dist) <- v.map$Group
colnames(v.dist) <- v.map$Group

b.dist <- vegan::vegdist(vegan::decostand(t(b.otu.ra), method = "hellinger"), method = "bray") %>% as.matrix()
b.map <- b.map[match(row.names(b.dist), b.map$SampleID),]
row.names(b.dist) <- b.map$Group
colnames(b.dist) <- b.map$Group

b.dist <- b.dist[match(row.names(v.dist), row.names(b.dist)), match(colnames(v.dist), colnames(b.dist))]
```

Run a Mantel test comparing viral vs microbial community structure
```{r}
vegan::mantel(as.dist(v.dist), as.dist(b.dist), permutations = 9999, method = "pearson") 
```
Calculate the tanglegram
```{r}
v.dd.row <- as.dendrogram(hclust(as.dist(v.dist)))
b.dd.row <- as.dendrogram(hclust(as.dist(b.dist)))

untang.dd <- dendlist(v.dd.row, b.dd.row) %>% 
  untangle() 
```

Plot viral dendogram
```{r}
v.ddata_x <- dendro_data(untang.dd[[1]])

v.p2 <- ggplot(segment(v.ddata_x)) +
  geom_segment(aes(x=x, y=y, xend=xend, yend=yend))

v.labs <- label(v.ddata_x) %>% 
  inner_join(v.map, by = c("label" = "Group")) 

v.dd <- v.p2 + 
  geom_point(data=v.labs,
             aes(x=x, y=0.05, fill = Month, shape = Month), size = 4, stroke = 1, color = "gray25") +
  coord_flip(expand = T) +
  scale_fill_manual(values = RColorBrewer::brewer.pal(6, "Paired")[3:4]) + 
  scale_shape_manual(values = c(21,24)) +
  theme_classic() +
  facet_wrap(~ "vOTUs\n(Virome)") +
  theme(text = element_text(size = 15),
        axis.text = element_blank(),
        axis.title = element_blank(),
        axis.line = element_blank(),
        axis.ticks = element_blank(),
        legend.position = "none") +
  scale_y_reverse()

v.dd
```
Plot microbial dendogram
```{r}
b.ddata_x <- dendro_data(untang.dd[[2]])

b.p2 <- ggplot(segment(b.ddata_x)) +
  geom_segment(aes(x=x, y=y, xend=xend, yend=yend))

b.labs <- label(b.ddata_x) %>% 
  inner_join(b.map, by = c("label" = "Group")) 

b.dd <- b.p2 + 
  geom_point(data=b.labs,
             aes(x=x, y=0.025, fill = Month, shape = Month), size = 4, stroke = 1, color = "gray25") +
  coord_flip(expand = T) +
  scale_fill_manual(values = RColorBrewer::brewer.pal(6, "Paired")[3:4]) + 
  scale_shape_manual(values = c(21,24)) +
  theme_classic() +
  facet_wrap(~ "16S OTUs\n(Total MG)") +
  theme(text = element_text(size = 15),
        axis.text = element_blank(),
        axis.title = element_blank(),
        axis.line = element_blank(),
        axis.ticks = element_blank(),
        legend.position = "none")

b.dd
```
Plot the tip connections and put it all together
```{r}
conx.p <- rbind(b.labs, v.labs) %>% 
  mutate(y = ifelse(Extraction == "BULK", 2, 1)) %>% 
  mutate(hjust = ifelse(Extraction == "BULK", 0, 1)) %>% 
  ggplot(aes(y,x)) +
  geom_text(aes(label = BchNtr, hjust = hjust, fill = Month)) +
  geom_line(aes(group = label, color = Month), size = 1) +
  scale_color_manual(values = RColorBrewer::brewer.pal(4, "Paired")[c(3,4)]) +
  scale_fill_manual(values = RColorBrewer::brewer.pal(4, "Paired")[c(3,4)]) +
  xlim(0.5,2.5) +
  theme_minimal() +
  theme(text = element_blank(),
                legend.position = "none",
        panel.grid = element_blank())

conx.p

dendro.p <- cowplot::plot_grid(v.dd, conx.p, b.dd, nrow = 1, align = "h", axis = "tb", rel_widths = c(2,3,2))
dendro.p
```

Plot the shifts in mean relative abundance for the set of differentially abundant OTUs across time points
```{r}
# Load DESeq2 results
v.dao <- readRDS("../Data/v_deseq_batch.RDS")
b.dao <- readRDS("../Data/16S_deseq_batch.RDS")

# Calculate the aggregated relative abundances for vOTUs
v.means <- v.dao %>% 
  mutate(TemporalEnrichment = fct_relevel(TemporalEnrichment, "NS")) %>% 
  inner_join(v.otu.tidy, by = "OTU_ID") %>% 
  group_by(SampleID, TemporalEnrichment) %>% 
  summarise(AggRelAb = sum(Count)) %>% 
  inner_join(v.map, by = "SampleID") %>% 
  group_by(Month, TemporalEnrichment) %>% 
  summarise(MeanRelAb = mean(AggRelAb)) %>% 
  mutate(Set = "vOTUs\n(Virome)")

# Calculate the aggregated relative abundances for 16S OTUs
b.means <- b.dao %>% 
  mutate(TemporalEnrichment = fct_relevel(TemporalEnrichment, "NS")) %>% 
  inner_join(b.otu.tidy, by = "OTU_ID") %>% 
  group_by(SampleID, TemporalEnrichment) %>% 
  summarise(AggRelAb = sum(Count)) %>% 
  inner_join(b.map, by = "SampleID") %>% 
  group_by(Month, TemporalEnrichment) %>% 
  summarise(MeanRelAb = mean(AggRelAb)) %>% 
  mutate(Set = "16S OTUs\n(Total MG)")

# Plot 
mean.ab.p <- rbind(v.means, b.means) %>% 
  mutate(Set = fct_relevel(Set, "vOTUs\n(Virome)")) %>% 
  ggplot(aes(Month, MeanRelAb, fill = TemporalEnrichment)) +
  geom_bar(stat = "identity", position = "fill") +
  scale_fill_manual(values = c("gray", RColorBrewer::brewer.pal(6, "Paired")[3:4]),
                    guide = guide_legend(title.hjust = 0.5,
                                         title.vjust = 0.5,
                                         title.position = "top")) +
  ylab("Mean Rel. Abund.") +
  facet_grid(. ~ Set) +
  theme_light() +
  theme(text = element_text(size = 15),
        axis.title.x = element_blank(),
        legend.position = "bottom",
        panel.border = element_blank())

mean.ab.p
```

Put it all together
```{r}
top <- cowplot::plot_grid(dendro.p, mean.ab.p + theme(legend.position = "none"), nrow = 1, rel_widths = c(8,6), labels = c("A", "B"), label_size = 20)
top
```

Add the nutrient heatmap. This part of the code needs the mean.ab.p and nut.heat.p plots (both coming from the nutrients.Rmd notebook) to be loaded in the environment
```{r}
#679:708
bottom <- cowplot::plot_grid(mean.ab.p, nut.heat.p, rel_widths = c(2,3))
bottom

cowplot::plot_grid(top, nut.heat.p, ncol = 1, rel_heights = c(9,11))
```

Get the taxonomic patterns of the set of differentially abundant vOTUs and 16S OTUs
```{r}
#Subset the significant OTUs
v.sig <- v.dao %>% 
  filter(p.adjusted < 0.05) 
b.sig <- b.dao %>% 
  filter(p.adjusted < 0.05)

# Plot viral taxonomy
a <- v.sig %>% 
  filter(Family != "Unassigned" & !is.na(Family) & Family != "Mixed") %>% 
  ggplot(aes(Family, fill = TemporalEnrichment)) +
  geom_bar() +
  scale_fill_manual(name = "Enrichment",
                    values = RColorBrewer::brewer.pal(4, "Paired")[3:4]) +
  ylab("Number of vOTUs") +
  xlab("Virus Family") +
  coord_flip() +
  theme_light() +
  theme(text = element_text(size = 15),
        legend.position = "top")

# Plot putative host taxonomy
b <- v.sig %>% 
  filter(HostPhylum != "Unassigned" & !is.na(HostPhylum) & HostPhylum != "Mixed") %>% 
  ggplot(aes(HostPhylum, fill = TemporalEnrichment)) +
  geom_bar() + 
  scale_fill_manual(values = RColorBrewer::brewer.pal(4, "Paired")[3:4]) +
  ylab("Number of vOTUs") +
  xlab("Host Phylum") +
  coord_flip() +
  theme_light() +
  theme(text = element_text(size = 15), 
        legend.position = "none")

# Plot 16S taxonomy
c <- b.sig %>% 
  mutate(PhyClass = fct_recode(PhyClass, 
                               "Unclassified Proteobacteria" = "unclass_Proteobacteria",
                               "Candidate Division WPS-1" = "candidate division WPS-1")) %>% 
  mutate(PhyClass = fct_relevel(PhyClass, "Unclassified Proteobacteria", "Gammaproteobacteria", "Deltaproteobacteria", "Betaproteobacteria", "Alphaproteobacteria")) %>% 
  ggplot(aes(PhyClass, fill = TemporalEnrichment)) +
  geom_bar() +
  scale_fill_manual(values = RColorBrewer::brewer.pal(4, "Paired")[3:4]) +
  ylab("Number of 16S OTUs") +
  xlab("Phylum / Proteobacteria Class") +
  coord_flip() +
  theme_light() +
  theme(text = element_text(size = 15),
        legend.position = "none")


# Put it all together
cowplot::plot_grid(a,b,c,
                   ncol = 1,
                   align = "v",
                   rel_heights = c(1,1,2),
                   labels = c("A", "B", "C"),
                   label_size = 20)
```