---
title: "votu_filtering"
output: html_notebook
---

This notebook formats the vOTU abundance tables to include only those vOTUs with >= 0.75 coverage

Load libraries
```{r}
library(tidyverse)
```

Get the name of the files
```{r}
bed.dir <- "../Data/Mapping/" 
bed.files <- list.files(path = bed.dir, pattern = "*bed.tsv", full.names = T)
bed.files

sampleid <- list.files(path = bed.dir, pattern = "*bed.tsv", full.names = F) %>% 
  str_remove(".vir.bed.tsv")
sampleid
```

Open the files
```{r}
bed.list <- lapply(bed.files, read.table, sep = "\t", header = F, col.names = c("contig", "start", "end", "coverage"))
names(bed.list) <- sampleid
```

Function to get coverage for each vOTUs in each bed file and filter out instances in which coverage is < 0.75
```{r}
get_coverage <- function(df) {
  df %>% 
    mutate(nbase = end - start) %>% 
    group_by(contig) %>% 
    mutate(length = sum(nbase)) %>% 
    mutate(perc_seq = nbase/length) %>% 
    filter(coverage > 0) %>% 
    summarise(total_coverage = sum(perc_seq)) %>% 
    filter(total_coverage >= 0.75)
}
```

Apply the function to all bed files
```{r}
cov.list <- lapply(bed.list, get_coverage)
cov.list
```

Make the list into a data frame
```{r}
cov.df <- plyr::ldply(cov.list, function(x) x) %>% 
  rename("SampleID" = ".id")
cov.df 
```

Convert it to a matrix
```{r}
cov.mtx <- cov.df %>% 
  spread(key = SampleID, value = total_coverage) %>% 
  as.data.frame() 

row.names(cov.mtx) <- cov.mtx$contig
cov.mtx <- cov.mtx[,-1]
cov.mtx
```

Remova NAs
```{r}
cov.mtx
cov.mtx <- !is.na(cov.mtx)
```

Get the tpmean table generated from BAMM and keep only those instances where coverage ≥ 0.75
```{r}
tpmean <- read.table("../Data/Mapping/biochar.vir.tpmean.tsv", header = T, sep = "\t", row.names = 1)
tpmean <- tpmean[,-1]
tpmean

tmp <- colnames(tpmean)
tmp  <-  str_remove(tmp, ".vir.sortedIndexed.bam")
colnames(tpmean) <- tmp

tpmean.75 <- tpmean[match(row.names(cov.mtx), row.names(tpmean)), match(colnames(cov.mtx), colnames(tpmean))]
tpmean.75 <- cov.mtx * tpmean.75
tpmean.75.tidy <- tpmean.75 %>% 
  as.data.frame() %>% 
  mutate(OTU_ID = row.names(.)) %>% 
  gather(key = "SampleID", value = "Abundance", -OTU_ID) %>% 
  select(SampleID, everything())
```

Get the count table generated from BAMM and keep only those instances where coverage ≥ 0.75
```{r}
count <- read.table("../Data/Mapping/biochar.vir.counts.tsv", header = T, sep = "\t", row.names = 1)
count <- count[,-1]
count

tmp <- colnames(count)
tmp  <-  str_remove(tmp, ".vir.sortedIndexed.bam")
colnames(count) <- tmp

count.75 <- count[match(row.names(cov.mtx), row.names(count)), match(colnames(cov.mtx), colnames(count))]
count.75 <- cov.mtx * count.75
count.75.tidy <- count.75 %>% 
  as.data.frame() %>% 
  mutate(OTU_ID = row.names(.)) %>% 
  gather(key = "SampleID", value = "Abundance", -OTU_ID) %>% 
  select(SampleID, everything())
```

Save formatted tables in the Analysis folder
```{r}
saveRDS(tpmean.75, "../../Analysis/Data/virome_tpm_75_mtx.RDS")
saveRDS(tpmean.75.tidy, "../../Analysis/Data/virome_tpm_75_tidy.RDS")

saveRDS(count.75, "../../Analysis/Data/virome_count_75_mtx.RDS")
saveRDS(count.75.tidy, "../../Analysis/Data/virome_count_75_tidy.RDS")
```

